---
title: "GSABenchmark"
author: "Andrei-Florian Stoica"
package: GSABenchmark
date: July 27, 2025
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{GSABenchmark}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

GSABenchmark is a package designed for benchmarking scRNA-seq gene 
set analysis methods. It provides both traditional and novel benchmark 
metrics, as well as visualization tools. Currently, GSABenchmark supports 
15 gene set analysis methods. 

# Installation
To install GSABenchmark, run the following commands in an R session:
```{r setup, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

BiocManager::install("andrei-stoica26/GSABenchmark")
```

# Prerequisites
In addition to GSABenchmark, you need to install 
[scRNAseq](https://bioconductor.org/packages/release/data/experiment/html/scRNAseq.html) 
and 
[scater](https://bioconductor.org/packages/release/bioc/html/scater.html) 
for this tutorial.

# Dataset loading

This tutorial uses an scRNA-seq human pancreas dataset with provided cell type 
annotations. 

After loading the required packages, download the dataset using the 
`BaronPancreasData` function from `scRNAseq`. The dataset will be stored as 
a `SingleCellExperiment` object. 

```{r message=FALSE, warning=FALSE, results=FALSE}
library(GSABenchmark)
library(scater)
library(scRNAseq)

scObj <- BaronPancreasData('human')

```

# Input preparation

GSABenchmark requires normalized and log-transformed data, and PCA coordinates.
We will use the functions `logNormCounts` from `scuttle` and runPCA 
from `scater` for these steps:


```{r}
scObj <- logNormCounts(scObj)
scObj <- runPCA(scObj)
```

# Creating gene sets

First, we will take a look at the cell types found in this dataset:

```{r}
table(colData(scObj)$label)
```

We will create gene sets for four cell types found in the data, using
cell type markers from [PanglaoDB](https://www.panglaodb.se/markers.html):

```{r}
acinarMarkers <- c('PRSS1', 'KLK1', 'CTRC', 'PNLIP', 'AKR1C3', 'CTRB1',
                   'DUOXA2', 'ALDOB', 'REG3A', 'SERPINA3', 'PRSS3', 'REG1B',
                   'CFB',	'GDF15',	'MUC1','ANPEP', 'ANGPTL4', 'OLFM4',
                   'GSTA1', 'LGALS2', 'PDZK1IP1', 'RARRES2', 'CXCL17',
                   'UBD', 'GSTA2', 'LYZ', 'RBPJL', 'PTF1A', 'CELA3A',
                   'SPINK1', 'ZG16', 'CEL', 'CELA2A', 'CPB1', 'CELA1',
                   'PNLIPRP1', 'RNASE1', 'AMY2B', 'CPA2','CPA1', 'CELA3B',
                   'CTRB2', 'PLA2G1B', 'PRSS2', 'CLPS', 'REG1A', 'SYCN')

alphaMarkers <- c('GCG', 'TTR', 'PCSK2', 'FXYD5', 'LDB2', 'MAFB',
                  'CHGA', 'SCGB2A1', 'GLS', 'FAP', 'DPP4', 'GPR119',
                  'PAX6', 'NEUROD1', 'LOXL4', 'PLCE1', 'GC', 'KLHL41',
                  'FEV', 'PTGER3', 'RFX6', 'SMARCA1', 'PGR', 'IRX1',
                  'UCP2', 'RGS4', 'KCNK16', 'GLP1R', 'ARX', 'POU3F4',
                  'RESP18', 'PYY', 'SLC38A5', 'TM4SF4', 'CRYBA2', 'SH3GL2',
                  'PCSK1', 'PRRG2', 'IRX2', 'ALDH1A1','PEMT', 'SMIM24',
                  'F10', 'SCGN', 'SLC30A8')

betaMarkers <- c('INS', 'IAPP', 'GJD2', 'PDX1', 'SLC2A2', 'NPY', 'MAFA', 
                 'PFKFB2','HOPX', 'PAX6', 'MAFB', 'CASR', 'EDARADD', 
                 'SCGB2A1', 'TGFBR3','ADCYAP1', 'SH3GL2', 'NEUROD1','ISL1', 
                 'RGS16', 'SMAD9', 'SIX3','BMP5', 'PIR', 'STXBP5', 'DLK1', 
                 'MEG3', 'GCGR', 'LMX1A', 'JPH3','CD40', 'HAMP','EZH1', 
                 'NTRK1','FXYD2', 'RIMS1', 'EFNA5', 'NPTX2','PAX4', 'PCSK2', 
                 'G6PC2', 'SLC30A8', 'PCSK1', 'SCGN', 'IGF2','SYT13', 'FFAR2', 
                 'SIX2')

deltaMarkers <- c('SST', 'GABRB3', 'FRZB', 'MS4A8', 'BAIAP3', 'CASR', 'BCHE',
                  'UNC5B','EDN3', 'GHSR', 'PCSK1', 'GABRG2', 'POU3F1',
                  'BHLHE41', 'EHF', 'LCORL', 'ETV1', 'PDX1', 'LEPR', 'UCP2',
                  'NPTX2', 'FXYD2', 'IAPP', 'KCNK16', 'SCGN', 'ISL1', 'HHEX',
                  'RESP18', 'PAX4', 'RBP4', 'PCSK9', 'FFAR4')
```

Now we gather the gene sets and we declare their names:

```{r}
geneSets <- list(acinarMarkers, alphaMarkers, betaMarkers, deltaMarkers)
geneSetNames <- c('acinar', 'alpha', 'beta', 'delta')
```

**!* Note: The gene set names must all be found in the column of assessed
class identities. For this dataset, this column is named `label`:

```{r}
setdiff(geneSetNames, colData(scObj)[['label']])
```

# Running the methods

The `supportedMethods` function lists all the methods currently available in 
`GSABenchmark`:

```{r}
supportedMethods()
```

For the purposes of this tutorial, we will select five of these methods:

```{r}
gsaMethods <- c('CSOA', 'PLAGE', 'VAM', 'SiPSiC', 'Zscore')
```

We now run the five methods for each gene set:

```{r}
scObj <- runGSAMethods(scObj, 'label', geneSets, gsaMethods, geneSetNames)
```

# Running the benchmark

The benchmark is run in a single line of code:

```{r}
smr <- runBenchmark(scObj, 'label', geneSets, gsaMethods, geneSetNames)
```

**!** Note: While this tutorial illustrates using `GSABenchmark` with 
Bioconductor's native `SingleCellExperiment` class, `GSABenchmark` is also fully
compatible with `Seurat`. Both `runGSAMethods` and `runBenchmark` also accept a
Seurat object as their first argument.

# Exploring the results

`runBenchmark` returns a list object with four elements. We can further explore
the structure of this object:

```{r}
names(smr)
names(smr$boundary)
names(smr$MCC)
names(smr$global)
names(smr$efficiency)
```

We can look at summaries in terms of correctness, speed, and memory usage:

```{r}
smr$boundary$metricSummary
smr$MCC
smr$global$metricSummary
smr$efficiency
```

To visualize the results, we will call the `allBenchmarkPlots` function. This
function plots each benchmark data frame:

```{r}
plots <- allBenchmarkPlots(smr)
length(plots)
```

Below, we display one of the resulting plots:

```{r, out.height='100%', out.width='100%', fig.height=6, fig.width=6}
plots[[3]]
```

# Session information {-}
```{r}
sessionInfo()
```
