---
title: "GSABenchmark"
author: "Andrei-Florian Stoica"
package: CSOA
date: July 27, 2025
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{GSABenchmark}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

GSABenchmark is a package designed for benchmarking scRNA-seq gene 
set analysis methods. It provides both traditional and novel benchmark 
metrics, as well as visualization tools. Currently, GSABenchmark supports 
15 gene set analysis methods. 

# Installation
To install GSABenchmark, run the following commands in an R session:
```{r setup, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

BiocManager::install("andrei-stoica26/GSABenchmark")
```

# Prerequisites
In addition to GSABenchmark, you need to install 
[scRNAseq](https://bioconductor.org/packages/release/data/experiment/html/scRNAseq.html) 
and 
[scuttle](https://www.bioconductor.org/packages/release/bioc/html/scuttle.html) 
for this tutorial.

# Dataset loading

This tutorial uses an scRNA-seq human pancreas dataset with provided cell type 
annotations. 

After loading the required packages, download the dataset using the 
`BaronPancreasData` function from `scRNAseq`. The dataset will be stored as 
a `SingleCellExperiment` object. 

```{r message=FALSE, warning=FALSE, results=FALSE}
library(ggplot2)
library(scRNAseq)
library(scuttle)
library(Seurat)

sceObj <- BaronPancreasData('human')

```

# Input preparation

GSABenchmark requires normalized and log-transformed data. We will employ the 
`logNormCounts` function from `scuttle` for this step:

```{r message=FALSE, warning=FALSE, results=FALSE}
sceObj <- logNormCounts(sceObj)
```

