---
title: "GSABenchmark"
author: "Andrei-Florian Stoica"
package: GSABenchmark
date: August 8, 2025
output: BiocStyle::html_document
vignette: >
    %\VignetteIndexEntry{GSABenchmark}
    %\VignetteEngine{knitr::rmarkdown}
    %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

GSABenchmark is a package designed for benchmarking scRNA-seq gene 
set analysis methods. It provides both traditional and novel benchmark 
metrics, as well as visualization tools. Currently, GSABenchmark supports 
15 gene set analysis methods. 

# Installation
To install GSABenchmark, run the following commands in an R session:
```{r setup, eval=FALSE}
if (!require("BiocManager", quietly=TRUE))
    install.packages("BiocManager")

BiocManager::install("andrei-stoica26/GSABenchmark")
```

# Prerequisites
In addition to GSABenchmark, you need to install 
[scRNAseq](https://bioconductor.org/packages/release/data/experiment/html/scRNAseq.html) 
and 
[scater](https://bioconductor.org/packages/release/bioc/html/scater.html) 
for this tutorial.

# Loading data

This tutorial uses an scRNA-seq human pancreas dataset with provided cell type 
annotations. 

After loading the required packages, download the dataset using the 
`BaronPancreasData` function from `scRNAseq`. The dataset will be stored as 
a `SingleCellExperiment` object. 

```{r message=FALSE, warning=FALSE, results=FALSE}
library(GSABenchmark)
library(scater)
library(scRNAseq)

scObj <- BaronPancreasData('human')

```

GSABenchmark requires normalized and log-transformed data, and PCA coordinates.
We will use the functions `logNormCounts` from `scuttle` and runPCA 
from `scater` for these steps:


```{r}
scObj <- logNormCounts(scObj)
scObj <- runPCA(scObj)
```

# Creating gene sets

First, we will take a look at the cell types found in this dataset:

```{r}
table(colData(scObj)$label)
```

We will create gene sets for four cell types found in the data, using
cell type markers from [PanglaoDB](https://www.panglaodb.se/markers.html):

```{r}
acinarMarkers <- c('PRSS1', 'KLK1', 'CTRC', 'PNLIP', 'AKR1C3', 'CTRB1',
                   'DUOXA2', 'ALDOB', 'REG3A', 'SERPINA3', 'PRSS3', 'REG1B',
                   'CFB',	'GDF15',	'MUC1','ANPEP', 'ANGPTL4', 'OLFM4',
                   'GSTA1', 'LGALS2', 'PDZK1IP1', 'RARRES2', 'CXCL17',
                   'UBD', 'GSTA2', 'LYZ', 'RBPJL', 'PTF1A', 'CELA3A',
                   'SPINK1', 'ZG16', 'CEL', 'CELA2A', 'CPB1', 'CELA1',
                   'PNLIPRP1', 'RNASE1', 'AMY2B', 'CPA2','CPA1', 'CELA3B',
                   'CTRB2', 'PLA2G1B', 'PRSS2', 'CLPS', 'REG1A', 'SYCN')

alphaMarkers <- c('GCG', 'TTR', 'PCSK2', 'FXYD5', 'LDB2', 'MAFB',
                  'CHGA', 'SCGB2A1', 'GLS', 'FAP', 'DPP4', 'GPR119',
                  'PAX6', 'NEUROD1', 'LOXL4', 'PLCE1', 'GC', 'KLHL41',
                  'FEV', 'PTGER3', 'RFX6', 'SMARCA1', 'PGR', 'IRX1',
                  'UCP2', 'RGS4', 'KCNK16', 'GLP1R', 'ARX', 'POU3F4',
                  'RESP18', 'PYY', 'SLC38A5', 'TM4SF4', 'CRYBA2', 'SH3GL2',
                  'PCSK1', 'PRRG2', 'IRX2', 'ALDH1A1','PEMT', 'SMIM24',
                  'F10', 'SCGN', 'SLC30A8')

betaMarkers <- c('INS', 'IAPP', 'GJD2', 'PDX1', 'SLC2A2', 'NPY', 'MAFA', 
                 'PFKFB2','HOPX', 'PAX6', 'MAFB', 'CASR', 'EDARADD', 
                 'SCGB2A1', 'TGFBR3','ADCYAP1', 'SH3GL2', 'NEUROD1','ISL1', 
                 'RGS16', 'SMAD9', 'SIX3','BMP5', 'PIR', 'STXBP5', 'DLK1', 
                 'MEG3', 'GCGR', 'LMX1A', 'JPH3','CD40', 'HAMP','EZH1', 
                 'NTRK1','FXYD2', 'RIMS1', 'EFNA5', 'NPTX2','PAX4', 'PCSK2', 
                 'G6PC2', 'SLC30A8', 'PCSK1', 'SCGN', 'IGF2','SYT13', 'FFAR2', 
                 'SIX2')

deltaMarkers <- c('SST', 'GABRB3', 'FRZB', 'MS4A8', 'BAIAP3', 'CASR', 'BCHE',
                  'UNC5B','EDN3', 'GHSR', 'PCSK1', 'GABRG2', 'POU3F1',
                  'BHLHE41', 'EHF', 'LCORL', 'ETV1', 'PDX1', 'LEPR', 'UCP2',
                  'NPTX2', 'FXYD2', 'IAPP', 'KCNK16', 'SCGN', 'ISL1', 'HHEX',
                  'RESP18', 'PAX4', 'RBP4', 'PCSK9', 'FFAR4')
```

Now we gather the gene sets and we declare their names:

```{r}
geneSets <- list(acinarMarkers, alphaMarkers, betaMarkers, deltaMarkers)
geneSetNames <- c('acinar', 'alpha', 'beta', 'delta')
```

**!* Note: The gene set names must all be found in the column of assessed
class identities. For this dataset, this column is named `label`:

```{r}
setdiff(geneSetNames, colData(scObj)[['label']])
```

# Running the methods

The `supportedMethods` function lists all the methods currently available in 
`GSABenchmark`:

```{r}
supportedMethods()
```

For the purposes of this tutorial, we will select 6 of these methods:

```{r}
gsaMethods <- c('AddModuleScore', 'CSOA', 'PLAGE', 'VAM', 'SiPSiC', 'Zscore')
```

We now run the five methods for each gene set:

```{r}
scObj <- runGSAMethods(scObj, 'label', geneSets, gsaMethods, geneSetNames)
```

# Running the benchmark

The benchmark is run in a single line of code. For the purposes of this 
vignette, we will perform only the correctness benchmark by setting 
`geneSets` to `NULL`. This command bypasses the execution of the speed and 
memory benchmarks.

```{r}
smr <- runBenchmark(scObj, 'label', geneSets=NULL, gsaMethods, geneSetNames)
```

**Note**: While this tutorial illustrates using `GSABenchmark` with 
Bioconductor's native `SingleCellExperiment` class, `GSABenchmark` is also fully
compatible with `Seurat`. Both `runGSAMethods` and `runBenchmark` also accept a
Seurat object as their first argument.

# Exploring the results

`runBenchmark` returns a list object with four elements. We can further explore
the structure of this object:

```{r}
names(smr)
names(smr$boundary)
names(smr$MCC)
names(smr$global)
names(smr$efficiency)
```

We can look at summaries in terms of correctness, speed, and memory usage:

```{r}
smr$boundary$metricSummary
smr$MCC
smr$global$metricSummary
smr$efficiency
```

To visualize the results, we will call the `allBenchmarkPlots` function. This
function plots each benchmark data frame:

```{r}
plots <- allBenchmarkPlots(smr)
length(plots)
```

Below, we display one of the resulting plots:

```{r, out.height='80%', out.width='80%', fig.height=5, fig.width=5}
plots[[3]]
```

This summary object is also used to create other plots. The `geneSetRankPlots` 
function creates plots ranking the gene set analysis methods for each gene set.
We display one of these plots:

```{r, out.height='80%', out.width='80%', fig.height=5, fig.width=5}
plots <- geneSetRankPlots(smr)
plots[[1]]
```

The `metricRankPlots` function creates plots ranking the gene set analysis 
methods for each benchmark metric. Below, we display one of these plots:

```{r, out.height='80%', out.width='80%', fig.height=5, fig.width=5}
plots <- metricRankPlots(smr)
plots[[5]]
```

We can also visualize the aggregate ranks of the methods using the 
`aggregateRankPlot` function, which also display the average rank obtained by
each method:

```{r, out.height='80%', out.width='80%', fig.height=5, fig.width=5}
aggregateRankPlot(smr)
```

Next, we will use the ratioRank function to visualize the top results among 
all method-metric-gene set combinations as ranked by the ratio of the 
maximum score and the mean:

```{r, out.height='100%', out.width='100%', fig.height=5, fig.width=5}
ratioPlot(smr, labelSize=2.5)
```

Finally, we will perform MDS plots for each gene set, illustrating how similar
or different the results obtained by different methods are. We showcase one of 
these plots below:

```{r, out.height='100%', out.width='100%', fig.height=5, fig.width=5}
plots <- mdsPlots(scObj, smr)
plots[[2]]
```

# Benchmarking a single method at different choices of gene loss and noise

GSABenchmark also offers the option of benchmarking a single method at 
different choice of gene loss (fraction of genes removed from each gene
signature) and noise (fraction of the retained genes in the final adjusted
gene signatures after random genes have been added).

Each condition determined by a gene loss-noise combination can be run in
replicates (that is, different gene signatures will be constructed with the
same gene loss and noise parameters for each replicate). The `seeds` argument
controls the random seeds used in the generation of gene sets, and its length
determines the number of replicates:

```{r}
geneSets <- list(alphaMarkers, betaMarkers)
geneSetNames <- c('alpha', 'beta')
scObj <- runMethodShuffle(scObj, 'label', geneSets, 'CSOA',
                              geneSetNames,
                              loss=c(1/3, 2/3),
                              noise=c(0.5),
                              seeds=c(20, 30))

```

Subsequently, the same benchmark as before can be run with 
`runBenchmarkShuffle`, a thin wrapper around `runBenchmark`:

```{r}
smr <- runBenchmarkShuffle(scObj, 'label', 'CSOA', geneSetNames)
```

We can then create the same plots as before before with the new benchmark
summary results. For instance, here we display a MDS plot:

```{r, out.height='100%', out.width='100%', fig.height=5, fig.width=5}
plots <- mdsPlots(scObj, smr)
plots[[2]]
```

# Session information {-}
```{r}
sessionInfo()
```
